Higher-Order unification for free!

### 2024-05-13 paper/appendix.tex
rename object constructors

  type tm->fm map -> ho.tm -> fo.fm -> o.
  tm->fm _ (ho.con C)  (fo.fcon C).
  tm->fm L (ho.lam B1) (fo.flam B2) :- 
    pi x y\ tm->fm _ x y => tm->fm L (B1 x) (B2 y).
  tm->fm L (ho.app L1) T :- map (tm->fm L) L1 [Hd|Tl], 
    fo.mk-app Hd Tl T.
  tm->fm L (ho.uva VM TL) T :- mem L (mapping (fv VO) (hv VM _)), 
    map (tm->fm L) TL T1, fo.mk-app (fo.fuva VO) T1 T.

>>>>

  type tm->fm map -> ho.tm -> fo.fm -> o.
  tm->fm _ (ho.con C)  (fo.o-con C).
  tm->fm L (ho.lam B1) (fo.o-lam B2) :- 
    pi x y\ tm->fm _ x y => tm->fm L (B1 x) (B2 y).
  tm->fm L (ho.app L1) T :- map (tm->fm L) L1 [Hd|Tl], 
    fo.mk-app Hd Tl T.
  tm->fm L (ho.uva VM TL) T :- mem L (mapping (fv VO) (hv VM _)), 
    map (tm->fm L) TL T1, fo.mk-app (fo.o-uva VO) T1 T.

### 2024-05-13 paper/main.tex
rename object constructors

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the ``arguments'' of a higher-order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.
In \Ho{} the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.
%
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity, we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak-head normal form of <<\elpiIn{fapp [F|L]}>>.

@@

%Applying the substitution corresponds to dereferencing a term with respect to
%the memory.
%  To ease the comparison we split \Fo{} dereferencing into a
% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
% to memory cells that are set with their values, and has a corresponding
% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
% has no corresponding operation in \Ho, and only ensures that
% terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
% <<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
% and \elpiIn{L2}. The reasons for this asymmetry is
% that an \elpiIn{fapp} node with a flexible head is always mapped
% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
% preventing nested applications to materialize.
% \input{code/fderef}
It is worth examining the code of \elpiIn{deref}, which
applies the substitution to a \Ho{} term. Notice how assignments are moved
to the current scope, i.e. the \elpiIn{abs}-bound variables are renamed
with the names in the scope of the unification variable occurrence.

@@

The final step is to decompile each assignment. The \elpiIn{val}
node carries a term that is easy to decompile since \mapStore is a bijection.
Since the \elpiIn{flam} node
carries no additional information (other than the function body),
each \elpiIn{abs} node can be decompiled to a \elpiIn{flam} one.

However, if the \elpiIn{flam} node was carrying more information,
as is the case for Coq where it holds the type of the bound variable,
one would need to store this piece of information in the memory map,
were we store the arity (that is a very simple function type).

@@

\begin{theorem}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{theorem}
\begin{proof}[Proof sketch]
In this setting, \Ee is as strong as \Eo on ground terms.
What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$,
we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, the only interesting case is
<<\elpiIn{fuva X ~\Uo~s}>>. In this case, after compilation, we have
<<\elpiIn{uva Y [] ~\Ue~t}>>
that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

  \begin{proof}[Proof sketch]
  If we look at the \Fo{} terms, there is one more interesting case,
  namely \elpiIn{fapp[fuva X|W] ~\Uo~s} when \elpiIn{W}
  are distinct names compiled to $\vec{w}$. In this case
  the \Ho problem is $Y_{\vec{w}} \Ue t$ that succeeds with
   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$, which in turn
   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.
   Thanks to \ref{clause:beta1}
   $(\lambda \vec{y}.s[\vec{w}/\vec{y}])~\vec{w} \Eo s$.
  \end{proof}

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
\end{lemma}
% \begin{proof}[Proof sketch]
%   problematic terms are mapped to uva by comp,
%   the problematic fapp node is gone.
%   if F x = t then  F = lam x.t[]
%   \todo{finish/maybe W enforcement is not needed here}
% \end{proof}
\noindent
In other word the compiler eliminates all subterms that are in \maybebeta.

@@

% \begin{elpicode}
% Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
%       , flam x\ fapp[fcon"g", fcon"a"] ]
% Problems = [ pr z (s z) ] % $\lambda x.g (F x) = \lambda x.g a$
% lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
% link z z (s z)
% HS = [some (abs x\con"a")]
% S = [some (flam x\fcon a)]
% \end{elpicode}

@@

% \begin{elpicode}
%   Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
%   , flam x\ fapp[fcon"g", fcon"a"] ]
% Problems = [ pr 0 1   % $A = \lambda x.x$
%            , pr 2 3 ] % $A a = a$
% lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
% link z z (s z)
% HS = [some (abs x\con"a")]
% S = [some (flam x\fcon a)]
% lam x\ app[f, app[X, x]] = Y,
%   lam x\ x) = X.
% \end{elpicode}

@@

\noindent
Whenever \elpiIn{flam F} is detected to be in
\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
variable \elpiIn{A}. This variable sees all the names free in
\elpiIn{lam F1} and is connected to \elpiIn{lam F1} via a \linketa.
\Fullref{inv:linklhs} holds for this link. Moreover:

@@

    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and, once the lambdas are crossed,
          $t_1$ and $t_2$ are unified. 
          This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.
  \end{itemize}
  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  % In all cases $X \Ue t2$ succeeds since X is a variable and
  % hence assigns t2 to X.
  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  % additional care of eta expanding t2 if needed (5.9.1).
% 
  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  % $s1$ with $s2$. The interesting case is when fostep succeds by
  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  % and that is in exactly the cases in which T (from page 3) is
  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
  % 
  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  % and the one flam-flam we find different terms.
  % (5.9.1) only adds lambdas that are forcibly consumed by
  % lam-lam rule.
\end{proof}

@@

% \begin{comment}
% \begin{proof}[Proof sketch]
%   Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
%   \Fo{} term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a
%   list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
%   however captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and
%   no \linkbeta is produced which contradicts our initial assumption.
% \end{proof}
% \end{comment}

@@

% \begin{comment}
% \begin{proof}[Proof sketch]
%   By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
%   the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
%   $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
%   in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
%   in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
%   \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
% \end{proof}
% \end{comment}

>>>>

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
Unification variables
in \Fo{} (\elpiIn{o-uva} term constructor) have no explicit scope:
the ``arguments'' of a higher-order variable are given via the \elpiIn{o-app}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{o-app [o-uva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.
In \Ho{} the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.
%
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity, we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak-head normal form of <<\elpiIn{o-app [F|L]}>>.

@@

%Applying the substitution corresponds to dereferencing a term with respect to
%the memory.
%  To ease the comparison we split \Fo{} dereferencing into a
% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
% to memory cells that are set with their values, and has a corresponding
% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
% has no corresponding operation in \Ho, and only ensures that
% terms of the form <<\elpiIn{o-app[o-app L1|L2]}>> are replaced by
% <<\elpiIn{o-app L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
% and \elpiIn{L2}. The reasons for this asymmetry is
% that an \elpiIn{o-app} node with a flexible head is always mapped
% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
% preventing nested applications to materialize.
% \input{code/fderef}
It is worth examining the code of \elpiIn{deref}, which
applies the substitution to a \Ho{} term. Notice how assignments are moved
to the current scope, i.e. the \elpiIn{abs}-bound variables are renamed
with the names in the scope of the unification variable occurrence.

@@

The final step is to decompile each assignment. The \elpiIn{val}
node carries a term that is easy to decompile since \mapStore is a bijection.
Since the \elpiIn{o-lam} node
carries no additional information (other than the function body),
each \elpiIn{abs} node can be decompiled to a \elpiIn{o-lam} one.

However, if the \elpiIn{o-lam} node was carrying more information,
as is the case for Coq where it holds the type of the bound variable,
one would need to store this piece of information in the memory map,
were we store the arity (that is a very simple function type).

@@

\begin{theorem}[Properties of \Uo in \wellb]\label{prop:uniffow}
  The implementation of \Uo above is a good unification
  for \Eo
  (as per \cref{prop:good-unif})
  in the domain \wellb.
\end{theorem}
\begin{proof}[Proof sketch]
In this setting, \Ee is as strong as \Eo on ground terms.
What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$,
we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, the only interesting case is
<<\elpiIn{o-uva X ~\Uo~s}>>. In this case, after compilation, we have
<<\elpiIn{uva Y [] ~\Ue~t}>>
that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

  \begin{proof}[Proof sketch]
  If we look at the \Fo{} terms, there is one more interesting case,
  namely \elpiIn{o-app[o-uva X|W] ~\Uo~s} when \elpiIn{W}
  are distinct names compiled to $\vec{w}$. In this case
  the \Ho problem is $Y_{\vec{w}} \Ue t$ that succeeds with
   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$, which in turn
   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.
   Thanks to \ref{clause:beta1}
   $(\lambda \vec{y}.s[\vec{w}/\vec{y}])~\vec{w} \Eo s$.
  \end{proof}

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
\end{lemma}
% \begin{proof}[Proof sketch]
%   problematic terms are mapped to uva by comp,
%   the problematic o-app node is gone.
%   if F x = t then  F = lam x.t[]
%   \todo{finish/maybe W enforcement is not needed here}
% \end{proof}
\noindent
In other word the compiler eliminates all subterms that are in \maybebeta.

@@

% \begin{elpicode}
% Terms [ o-lam x\ o-app[o-con"g",o-app[o-uva z, x]]
%       , o-lam x\ o-app[o-con"g", o-con"a"] ]
% Problems = [ pr z (s z) ] % $\lambda x.g (F x) = \lambda x.g a$
% lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
% link z z (s z)
% HS = [some (abs x\con"a")]
% S = [some (o-lam x\o-con a)]
% \end{elpicode}

@@

% \begin{elpicode}
%   Terms [ o-lam x\ o-app[o-con"g",o-app[o-uva z, x]]
%   , o-lam x\ o-app[o-con"g", o-con"a"] ]
% Problems = [ pr 0 1   % $A = \lambda x.x$
%            , pr 2 3 ] % $A a = a$
% lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
% link z z (s z)
% HS = [some (abs x\con"a")]
% S = [some (o-lam x\o-con a)]
% lam x\ app[f, app[X, x]] = Y,
%   lam x\ x) = X.
% \end{elpicode}

@@

\noindent
Whenever \elpiIn{o-lam F} is detected to be in
\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
variable \elpiIn{A}. This variable sees all the names free in
\elpiIn{lam F1} and is connected to \elpiIn{lam F1} via a \linketa.
\Fullref{inv:linklhs} holds for this link. Moreover:

@@

    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and, once the lambdas are crossed,
          $t_1$ and $t_2$ are unified. 
          This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.
  \end{itemize}
  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  % In all cases $X \Ue t2$ succeeds since X is a variable and
  % hence assigns t2 to X.
  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  % additional care of eta expanding t2 if needed (5.9.1).
% 
  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  % $s1$ with $s2$. The interesting case is when fostep succeds by
  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  % and that is in exactly the cases in which T (from page 3) is
  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
  % 
  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  % and the one o-lam-o-lam we find different terms.
  % (5.9.1) only adds lambdas that are forcibly consumed by
  % lam-lam rule.
\end{proof}

@@

% \begin{comment}
% \begin{proof}[Proof sketch]
%   Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
%   \Fo{} term has the shape \elpiIn{o-app[o-uva M | Ag]} where \elpiIn{Ag} is a
%   list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
%   however captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and
%   no \linkbeta is produced which contradicts our initial assumption.
% \end{proof}
% \end{comment}

@@

% \begin{comment}
% \begin{proof}[Proof sketch]
%   By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
%   the list \elpiIn{Ag} from the original term \elpiIn{o-app [o-uva A|Ag]}.
%   $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
%   in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
%   in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
%   \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
% \end{proof}
% \end{comment}
